---
title: "Getting started with poEvaluationMetrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with poEvaluationMetrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package evaluates species distribution models (SDMs) with discrimination metrics (AUC, correlation, PRG), calibration error (MAE, bias), Boyce-type indices (`SBI_*`), and threshold-based skill scores (TSS, Kappa, `Fbp`, SEDI, ORSS, omission, and related helpers).

## Standalone metric helpers

Many helpers take **binary** `actual` and `predicted` vectors (`0`/`1`):

```{r helpers}
library(poEvaluationMetrics)

actual <- c(1, 1, 0, 0)
pred_bin <- c(1, 0, 0, 1)

Fbp(actual, pred_bin)
omission(actual, pred_bin)
orss(actual, pred_bin)
brier_score(c(0, 1, 1, 0), c(0.2, 0.7, 0.9, 0.1))
```

## From probabilities to a full metric table

`indexCalculation()` expects a `data.frame` with columns `predicted` (numeric scores) and `observed` (`0`/`1`), plus a `SpatRaster` of predictions (used for Boyce/SBI sampling and related steps).

```{r indexCalculation, message = FALSE}
library(terra)
library(sf)

set.seed(1)
r <- rast(
  nrows = 80,
  ncols = 80,
  xmin = 0,
  xmax = 1,
  ymin = 0,
  ymax = 1,
  crs = "local"
)
values(r) <- runif(ncell(r))

inputDF <- data.frame(
  predicted = c(0.85, 0.72, 0.25, 0.41, 0.05, 0.91),
  observed = c(1L, 1L, 0L, 0L, 0L, 1L)
)

round(indexCalculation(inputDF, prediction = r), 4)
```

## Presence and background as `sf` points

`calculateMetrics()` extracts values at presence and absence/background points, then calls `indexCalculation()`.

```{r calculateMetrics}
pres_coords <- data.frame(x = c(0.2, 0.35, 0.6), y = c(0.25, 0.7, 0.45))
bg_coords   <- data.frame(x = c(0.1, 0.5, 0.85), y = c(0.55, 0.15, 0.75))

presence <- st_as_sf(pres_coords, coords = c("x", "y"), crs = crs(r))
absence  <- st_as_sf(bg_coords, coords = c("x", "y"), crs = crs(r))

round(calculateMetrics(r, presence, absence), 4)
```

## Multiple evaluation designs with `performanceEstimation()`

`performanceEstimation()` combines presence–absence, presence–background (replicated background sampling), and/or presence–artificial-absence (AOA) evaluation. It returns one **`data.frame`** with a **`scenario`** column (`PA`, `PBG`, `PAA`).

The chunk below keeps replication small for illustration. Turning on `aa = TRUE` uses **CAST** and is slower; here we disable PAA and use only PBG so the vignette builds quickly.

```{r performanceEstimation, message = FALSE}
env <- c(r, r * 0.8 + 0.1)
names(env) <- c("v1", "v2")

pe_result <- performanceEstimation(
  prediction = r,
  presence = presence,
  absence = absence,
  background = TRUE,
  aa = FALSE,
  environmentalVariables = env,
  noPointsTesting = 20,
  replicates = 2
)

pe_result[, 1:8]
```

See `?performanceEstimation` for all arguments and design options.

## Further reading

- Function reference in `help(package = "poEvaluationMetrics")`.
- README dependency notes for **prg** and **mecofun** if you install from Git.
